---
title: "End-to-End Predictive Modeling: Steam Games"
author: "Group 15: Buyang Daffa, Cathy Zhang, Elaine Tao, Thê Quach"
date: today
format:
  html:
    toc: true
bibliography: references.bib
editor: source
---

```{r}
#| message: false
#| warning: false

required_packages <- c(
  "jsonlite", "tidyverse", "lubridate", "caret", "scales",
  "patchwork", "purrr", "janitor", "pROC"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(paste0(
    "Please install these packages before running the notebook: ",
    paste(missing_packages, collapse = ", ")
  ))
}

invisible(lapply(required_packages, library, character.only = TRUE))

set.seed(123)
```

# Summary

This report presents a full end-to-end predictive modeling workflow for Steam game
metadata, aimed at predicting whether a game is free-to-play or paid. Using a snapshot
of approximately 5,000 Steam games collected in August 2025 (@vintagedon2025dataset),
we engineer features from game attributes — including genre, supported platforms,
developer, and release year — and train a cross-validated logistic regression classifier.
The final model achieves 95.9% accuracy and a ROC AUC of 0.923 on the held-out test
set, substantially outperforming a majority-class baseline of 79.4%.

# Introduction

Steam is the world's largest PC game distribution platform, with over 10,000 annual
releases (@statista2026steam) and around 20 million daily active users (@steam_charts).
Its open submission program allows indie and small studios to reach a global audience,
making the platform's catalogue highly diverse in both genre and monetisation strategy.

Being able to predict whether a game will be free or not will help groups determine
pricing structure, target audience, and game longevity. The following groups would be
particularly interested:

- Developers making pre-launch decisions about pricing, feature scope, and genre
  positioning.
- Publishers evaluating which projects to greenlight or invest in.
- Platform researchers studying what drives quality signals in large content marketplaces.

The dataset used is the Steam 2025 5K Games Dataset (@vintagedon2025dataset):
a gzip-compressed JSON snapshot of approximately 5,000 Steam game records taken on
August 31, 2025. Key fields include `steam_appid`, `name`, `is_free` (the target
variable), `genres`, `categories`, `platforms`, `release_date`, `required_age`, and
`developer`.

Specifically, we ask:

1. What game attributes differ most between free-to-play and paid titles?
2. Can a machine learning classifier accurately distinguish free and paid games using
   only pre-launch observable features?
3. Which features contribute most to that classification?

# Methods & Results

## Data Loading and Cleaning

The R programming language (@R) and the following R packages were used to perform
the analysis: tidyverse (@tidyverse), caret (@caret), pROC (@pROC), janitor
(@janitor), lubridate (@lubridate), scales (@scales), patchwork (@patchwork),
jsonlite (@jsonlite), and Quarto (@Allaire_Quarto_2022).

The dataset is loaded from a local JSON file. Multiple candidate paths are checked
so the script runs across different directory layouts. Column names are simplified
from verbose dot-separated API paths to concise labels, and key type conversions
are applied: `is_free` becomes a two-level factor (`"Free"` / `"Paid"`), platform
flags are cast to logical and summed into `platform_count`, and release year is
extracted from raw date strings using `lubridate::mdy()`. The top 20 most-frequent
game categories are one-hot encoded as `cat_*` indicator columns. Missing values
are imputed with sensible defaults rather than dropped, keeping data loss below 10%.
Duplicate `game_id` rows are removed with `distinct()`.

```{r}
#| message: false
#| warning: false

# ---- Locate and load the dataset ----
possible_paths <- c(
  "../data/games_sample.json",
  "data/games_sample.json",
  "./data/games_sample.json",
  "/mnt/data/games_sample.json"
)

data_path <- possible_paths[file.exists(possible_paths)][1]

if (is.na(data_path) || length(data_path) == 0) {
  stop(paste(
    "Could not find games_sample.json. Expected one of:",
    paste(possible_paths, collapse = ", ")
  ))
}

raw <- fromJSON(data_path, flatten = TRUE)

if (!"games" %in% names(raw)) {
  stop("The JSON file does not contain a top-level `games` field.")
}

df <- raw$games

if ("app_details.data.categories" %in% names(df)) {
  attr(df[["app_details.data.categories"]], "dim") <- NULL
}

# ---- Feature engineering ----
flat_data <- df |>
  rename(
    game_id         = appid,
    game_name       = name_from_applist,
    required_age    = app_details.data.required_age,
    is_free         = app_details.data.is_free,
    game_type       = app_details.data.type,
    release_date    = app_details.data.release_date.date,
    windows_support = app_details.data.platforms.windows,
    mac_support     = app_details.data.platforms.mac,
    linux_support   = app_details.data.platforms.linux,
    categories      = app_details.data.categories,
    dlc             = app_details.data.dlc,
    demos           = app_details.data.demos,
    developers      = app_details.data.developers,
    publishers      = app_details.data.publishers
  ) |>
  mutate(
    is_free             = factor(is_free, levels = c(TRUE, FALSE), labels = c("Free", "Paid")),
    required_age        = suppressWarnings(as.integer(required_age)),
    game_type           = as.factor(game_type),
    release_date_parsed = suppressWarnings(mdy(release_date)),
    release_year        = year(release_date_parsed),
    windows_support     = as.logical(windows_support),
    mac_support         = as.logical(mac_support),
    linux_support       = as.logical(linux_support),
    platform_count      = as.integer(replace_na(windows_support, FALSE)) +
                          as.integer(replace_na(mac_support,     FALSE)) +
                          as.integer(replace_na(linux_support,   FALSE)),
    has_dlc             = map_lgl(dlc,   ~ length(.x) > 0),
    has_demo            = map_lgl(demos, ~ length(.x) > 0),
    category_list       = map(categories, function(cat) {
      if (is.null(cat) || !is.data.frame(cat) || nrow(cat) == 0) return(character(0))
      if ("description" %in% names(cat)) return(cat$description)
      character(0)
    }),
    n_categories   = map_int(category_list, length),
    developer_name = map_chr(developers, ~ if (length(.x) > 0) .x[[1]] else NA_character_),
    publisher_name = map_chr(publishers, ~ if (length(.x) > 0) .x[[1]] else NA_character_)
  )

all_cats <- flat_data |>
  select(game_id, category_list) |>
  unnest(category_list) |>
  rename(category = category_list)

top_categories <- all_cats |>
  count(category, sort = TRUE) |>
  slice_head(n = 20) |>
  pull(category)

for (cat_name in top_categories) {
  col_name <- paste0("cat_", janitor::make_clean_names(cat_name))
  flat_data[[col_name]] <- map_lgl(flat_data$category_list, ~ cat_name %in% .x)
}

cat_cols <- grep("^cat_", names(flat_data), value = TRUE)

df_model <- flat_data |>
  transmute(
    game_id,
    game_name,
    is_free,
    required_age    = replace_na(required_age, 0L),
    release_year    = replace_na(release_year, -1L),
    game_type       = fct_na_value_to_level(game_type, level = "unknown"),
    windows_support = replace_na(windows_support, FALSE),
    mac_support     = replace_na(mac_support,     FALSE),
    linux_support   = replace_na(linux_support,   FALSE),
    platform_count,
    has_dlc         = replace_na(has_dlc,  FALSE),
    has_demo        = replace_na(has_demo, FALSE),
    n_categories    = replace_na(n_categories, 0L),
    developer_name,
    publisher_name,
    across(all_of(cat_cols), ~ replace_na(.x, FALSE))
  ) |>
  filter(!is.na(is_free)) |>
  distinct(game_id, .keep_all = TRUE)
```

## Exploratory Data Analysis

The dataset is heavily imbalanced: paid games account for 79.4% of records and free
games for only 20.6% (@fig-target-dist). This means a naive majority-class classifier
would already achieve ~79% accuracy, so model accuracy must be interpreted alongside
sensitivity and specificity.

```{r}
#| label: fig-target-dist
#| fig-cap: "Class distribution of the target variable. Paid games outnumber free games roughly 4:1."
#| fig-width: 5
#| fig-height: 4

class_counts <- df_model |>
  count(is_free) |>
  mutate(
    pct   = n / sum(n),
    label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")
  )

ggplot(class_counts, aes(x = is_free, y = n, fill = is_free)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, size = 4.3, fontface = "bold") +
  scale_fill_manual(values = c("Free" = "#10B981", "Paid" = "#3B82F6")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Target Distribution: Free vs Paid Games", x = NULL, y = "Number of Games") +
  theme_minimal()
```

@fig-numeric-dist shows distributions of the four key numeric predictors. `n_categories`
is right-skewed: most games belong to just a few categories. `platform_count` clusters
at 1, reflecting that most titles are Windows-only. `release_year` rises sharply
post-2015 as Steam's catalogue expanded. `required_age` is concentrated at 0, as most
games carry no age restriction.

```{r}
#| label: fig-numeric-dist
#| fig-cap: "Distributions of the four key numeric predictors."

numeric_vars <- df_model |>
  select(required_age, release_year, platform_count, n_categories) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  drop_na()

ggplot(numeric_vars, aes(x = value, fill = variable)) +
  geom_histogram(bins = 40, color = "white", linewidth = 0.2, show.legend = FALSE) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Distribution of Key Numeric Predictors", x = NULL, y = "Count") +
  theme_minimal()
```

@fig-feature-rel presents three complementary views of feature relationships by class.
Release-year distributions are similarly shaped for both classes, but paid games
dominate numerically throughout. The binary feature panel is the most striking: paid
games are far more likely to have DLC or a demo available, consistent with using demos
to drive paid conversions.

```{r}
#| label: fig-feature-rel
#| fig-cap: "Feature relationships broken down by free vs paid status."
#| fig-height: 14

p1 <- ggplot(df_model |> filter(release_year > 0), aes(x = release_year, fill = is_free)) +
  geom_histogram(position = "identity", alpha = 0.55, bins = 30) +
  scale_fill_manual(values = c("Free" = "#10B981", "Paid" = "#3B82F6")) +
  labs(title = "Release Year by Class", x = "Release Year", y = "Count", fill = NULL) +
  theme_minimal()

p2 <- df_model |>
  mutate(n_categories_capped = pmin(n_categories, 15L)) |>
  ggplot(aes(x = factor(n_categories_capped), y = platform_count, fill = is_free)) +
  geom_boxplot(alpha = 0.65, outlier.size = 0.7) +
  scale_fill_manual(values = c("Free" = "#10B981", "Paid" = "#3B82F6")) +
  labs(
    title = "Platform Count by Number of Categories",
    x = "Number of Categories (capped at 15)", y = "Platform Count", fill = NULL
  ) +
  theme_minimal()

p3 <- df_model |>
  group_by(is_free) |>
  summarise(`Has DLC` = mean(has_dlc), `Has Demo` = mean(has_demo), .groups = "drop") |>
  pivot_longer(-is_free, names_to = "feature", values_to = "pct") |>
  ggplot(aes(x = feature, y = pct, fill = is_free)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.6), vjust = -0.35, size = 3.8
  ) +
  scale_fill_manual(values = c("Free" = "#10B981", "Paid" = "#3B82F6")) +
  scale_y_continuous(labels = percent, limits = c(0, 1.05)) +
  labs(title = "Binary Feature Rates by Class", x = NULL, y = "Proportion", fill = NULL) +
  theme_minimal()

p1 / p2 / p3
```

@fig-cat-analysis summarises category-level patterns. Single-player is the most
prevalent tag overall, followed by family sharing and Steam achievements. The divergence
panel shows that single-player games strongly skew paid, while multiplayer and Steam
leaderboard games tilt slightly toward free — consistent with free-to-play titles
relying on large player pools for engagement.

```{r}
#| label: fig-cat-analysis
#| fig-cap: "Category coverage, prevalence, and divergence between free and paid games."
#| fig-height: 18

cat_coverage <- df_model |>
  mutate(has_categories = n_categories > 0) |>
  group_by(is_free) |>
  summarise(pct = mean(has_categories), .groups = "drop")

p1 <- ggplot(cat_coverage, aes(x = is_free, y = pct, fill = is_free)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), vjust = -0.4, size = 4.4, fontface = "bold") +
  scale_fill_manual(values = c("Free" = "#10B981", "Paid" = "#3B82F6")) +
  scale_y_continuous(labels = percent, limits = c(0, 1.05)) +
  labs(title = "% of Games with At Least One Category", x = NULL, y = "% with Categories") +
  theme_minimal()

p2 <- all_cats |>
  count(category, sort = TRUE) |>
  slice_head(n = 20) |>
  mutate(pct = n / nrow(df_model)) |>
  ggplot(aes(x = reorder(category, pct), y = pct)) +
  geom_col(fill = "#6366F1", width = 0.7) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = percent, limits = c(0, 1.05)) +
  labs(title = "Top 20 Categories by Prevalence", x = NULL, y = "% of All Games") +
  theme_minimal()

cat_prev <- flat_data |>
  select(game_id, is_free, category_list) |>
  unnest(category_list) |>
  rename(category = category_list) |>
  filter(category %in% top_categories) |>
  count(is_free, category) |>
  left_join(df_model |> count(is_free, name = "group_total"), by = "is_free") |>
  mutate(pct = n / group_total)

cat_diff <- cat_prev |>
  select(is_free, category, pct) |>
  pivot_wider(names_from = is_free, values_from = pct, values_fill = 0) |>
  mutate(diff = Free - Paid, direction = if_else(diff > 0, "Free > Paid", "Paid > Free")) |>
  arrange(desc(abs(diff)))

p3 <- ggplot(cat_diff, aes(x = reorder(category, diff), y = diff, fill = direction)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = paste0(if_else(diff > 0, "+", ""), percent(diff, accuracy = 0.1))),
    hjust = if_else(cat_diff$diff > 0, -0.1, 1.1), size = 3.1
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Free > Paid" = "#10B981", "Paid > Free" = "#3B82F6")) +
  scale_y_continuous(labels = percent, limits = c(-0.85, 0.85)) +
  labs(
    title = "Difference in Category Prevalence (Free - Paid)",
    x = NULL, y = "Difference in Prevalence", fill = NULL
  ) +
  theme_minimal()

p1 / p2 / p3
```

## Modeling

The dataset is split 80/20 into training and test sets using stratified sampling
(`createDataPartition`) to preserve the class ratio. Only features observable before
a game's commercial release are included as predictors, making the model applicable to
pre-launch decisions. A majority-class baseline — predicting "Paid" for every
observation — yields 79.4% accuracy, the benchmark our model must beat.

Logistic regression is chosen for its interpretability and ease of deployment. We use
`caret::train()` (@caret) with 5-fold cross-validation, optimising for ROC AUC.

```{r}
#| message: false
#| warning: false

set.seed(123)
train_idx <- createDataPartition(df_model$is_free, p = 0.80, list = FALSE)
train_df  <- df_model[train_idx, ]
test_df   <- df_model[-train_idx, ]

predictor_names <- c(
  "required_age", "release_year", "game_type",
  "windows_support", "mac_support", "linux_support",
  "platform_count", "has_dlc", "has_demo", "n_categories",
  grep("^cat_", names(df_model), value = TRUE)
)

train_model <- train_df |> select(is_free, all_of(predictor_names))
test_model  <- test_df  |> select(is_free, all_of(predictor_names))

baseline_class    <- train_model |> count(is_free, sort = TRUE) |> slice(1) |> pull(is_free) |> as.character()
baseline_pred     <- factor(rep(baseline_class, nrow(test_model)), levels = levels(test_model$is_free))
baseline_cm       <- confusionMatrix(baseline_pred, test_model$is_free, positive = "Free")
baseline_accuracy <- unname(baseline_cm$overall["Accuracy"])

ctrl <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE, summaryFunction = twoClassSummary,
  savePredictions = "final"
)

model_formula <- as.formula(paste("is_free ~", paste(predictor_names, collapse = " + ")))

set.seed(123)
fit_glm <- caret::train(
  model_formula, data = train_model,
  method = "glm", family = binomial(),
  metric = "ROC", trControl = ctrl
)

fit_glm
```

## Evaluation

We evaluate the model on the held-out test set using probability-based predictions
thresholded at 0.5. We present classification metrics in @tbl-metrics.

```{r}
#| label: tbl-metrics
#| tbl-cap: "Classification metrics for the logistic regression model on the held-out test set."
#| message: false

prob_free  <- predict(fit_glm, newdata = test_model, type = "prob")$Free
pred_class <- factor(ifelse(prob_free >= 0.5, "Free", "Paid"), levels = levels(test_model$is_free))

cm <- confusionMatrix(pred_class, test_model$is_free, positive = "Free")
roc_obj <- pROC::roc(
  response  = test_model$is_free,
  predictor = prob_free,
  levels    = c("Paid", "Free"),
  direction = "<"
)

metrics_tbl <- tibble(
  Metric = c(
    "Baseline Accuracy", "Model Accuracy",
    "Sensitivity (Recall for Free)", "Specificity",
    "Balanced Accuracy", "ROC AUC"
  ),
  Value = round(c(
    baseline_accuracy,
    unname(cm$overall["Accuracy"]),
    unname(cm$byClass["Sensitivity"]),
    unname(cm$byClass["Specificity"]),
    unname(cm$byClass["Balanced Accuracy"]),
    as.numeric(pROC::auc(roc_obj))
  ), 4)
)

knitr::kable(metrics_tbl)
```

The model improves on the baseline by 16 percentage points. Specificity is near-perfect
at 0.990, while sensitivity reaches 0.838. Cohen's Kappa of 0.867 confirms performance
well beyond chance (@cadili2015cohens). Balanced accuracy of 0.914 — the mean per-class
recall — indicates the model handles both classes reasonably despite the imbalance
(@desilva2024balanced).

We can see from @tbl-metrics that the model achieves an overall accuracy of 95.9%.
In @fig-roc we show the ROC curve, which hugs the top-left corner (AUC = 0.923),
confirming strong discriminative ability.

```{r}
#| label: fig-roc
#| fig-cap: "ROC curve on the held-out test set. AUC = 0.923."
#| fig-width: 5
#| fig-height: 5

plot(roc_obj, main = "Test ROC Curve: Predicting Free vs Paid")
abline(a = 0, b = 1, lty = 2, col = "gray50")
```

The five most important predictors identified by the model are family sharing,
single-player, Steam leaderboards, multiplayer, and demo availability. Variable
importance scores are presented in @tbl-varimp.

```{r}
#| label: tbl-varimp
#| tbl-cap: "Top 15 variable importance scores from the logistic regression model."

coef_tbl <- summary(fit_glm$finalModel)$coefficients |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  as_tibble() |>
  arrange(desc(abs(Estimate))) |>
  slice_head(n = 15) |>
  select(Term, Estimate, `Std. Error`, `z value`, `Pr(>|z|)`) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

knitr::kable(coef_tbl)
```

Note that we define standard deviation (of a sample) as

$$s = \sqrt{\frac{\sum_{i=1}^N (x_i - \overline{x})^2}{N-1}}$$

and that family sharing (`cat_family_sharing`) has the largest absolute coefficient
of all predictors in @tbl-varimp, making it the single most influential feature in
the model.

# Discussion

The logistic regression classifier achieves strong overall performance (accuracy 0.959,
AUC 0.923), but the asymmetry between sensitivity (0.838) and specificity (0.990)
warrants careful interpretation. The model is more reliable at identifying paid games
than free ones — a consequence of the 4:1 class imbalance in the training data. For
applications where correctly flagging free games matters most, a lower decision
threshold or class-weighting during training would be preferable.

The variable importance results are counterintuitive in places. Family sharing emerges
as the single strongest predictor of a game being *free*, yet the EDA showed that paid
games have family sharing more often in absolute frequency. The logistic regression
coefficient captures the conditional relationship after controlling for all other
variables, which can reverse marginal associations — a known consequence of
multivariate adjustment.

The strong influence of the single-player category on predicting *paid* games aligns
with industry convention: narrative or campaign-driven titles typically require a
purchase, whereas free-to-play games rely on multiplayer engagement and microtransactions
for revenue. Similarly, the presence of Steam leaderboards as a predictor for free games
is consistent with competitive, score-driven free-to-play titles.

An interesting follow-up analysis would be to define multiplayer sub-categories
(MOBA, MMORPG, co-op) to determine whether finer genre distinctions further improve
predictive power, or to extend the target variable beyond a binary free/paid split
to capture the full pricing distribution.

# References
