---
title: "What makes a good game?"
author: "Data Analysis Project"
date: "2026-03-01`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)
```

------------------------------------------------------------------------

# Summary

This project aims to perform ananalysis of Steam game success, defined as achieving a positive review ratio above 80% ("Very Positive" or better on Steam's scale). Using a dataset of \~5,000 Steam games collected in August 2025, we will use game's attributes to train classification models to predict whether a game will be successful or not based on the following attributes: price, genre, supported platform features, developer, release year, and free-to-play status.

------------------------------------------------------------------------

# 1. Introduction

## 1.1 Background

Steam is the world's largest PC game distribution platform with over 70,000 titles and millions of daily active users as of 2025. Developers can publish games through an open submission program letting indie and small creators show off their games to a wider audience.

User reviews allow Steam's recommendation system to suggest users with potential titles they'd be interested in.

Being able to predict whether a game's design and feature choices are likely to earn strong reviews would be valuable to:

-   **Developers** making pre-launch decisions about pricing, feature scope, and genre positioning
-   **Publishers** evaluating which projects to greenlight
-   **Platform researchers** studying what drives quality signals in large content marketplaces

## 1.2 Research Question

Can we predict whether a Steam game will achieve a "Very Positive" or better review rating using observable game attributes available at or near launch?

Specifically, we ask:

1.  What specifically are players interested in when looking at games? (price tier, genre, platform features, free-to-play status)
2.  Can machine learning classifiers accurately distinguish high-rated from lower-rated games?
3.  Aside from actual quality of the game, what do users look for in games?

## 1.3 Dataset Description

The dataset is the **Steam 2025 5K Games Dataset**:

> **Source:** [`steam_2025_5k-dataset-games_20250831.json.gz`](https://github.com/vintagedon/steam-dataset-2025/blob/main/data/01_raw/steam_2025_5k-dataset-games_20250831.json.gz)\
> **Maintainer:** vintagedon (GitHub)\
> **Snapshot date:** August 31, 2025\
> **Format:** Gzip-compressed JSON\
> **Size:** \~5,000 game records

Key fields used in this analysis:

| Field | Type | Description |
|------------------------|------------------------|------------------------|
| `steam_appid` | integer | Unique game identifier |
| `name` | string | Game title |
| `is_free` | boolean | Free-to-play status |
| `price_overview` | nested dict | Current price in USD |
| `genres` | list of dicts | Official Steam genres |
| `categories` | list of dicts | Feature flags (multiplayer, achievements, etc.) |
| `tags` | dict / list | Community-applied descriptive labels |
| `total_positive` | integer | Number of positive reviews |
| `total_negative` | integer | Number of negative reviews |
| `platforms` | dict | Supported OS (Windows, Mac, Linux) |
| `release_date` | nested dict | Release date string |
| `required_age` | integer | Age rating |
| `developer` | list of char | Developer studio/company |

------------------------------------------------------------------------

# 2. Setup

```{r libraries}
library(tidyverse)  
library(jsonlite)   
library(lubridate)       
library(vip) 
library(scales)     
library(ggcorrplot)      
library(patchwork)

theme_set(theme_minimal(base_size = 12))

cat("All libraries loaded.\n")
```

------------------------------------------------------------------------

# 3. Data Loading & Inspection

```{r load-data}
DATA_URL <- "https://github.com/vintagedon/steam-dataset-2025/raw/refs/heads/main/data/01_raw/steam_2025_5k-dataset-games_20250831.json.gz"

steam_data <- fromJSON(
  gzcon(url(DATA_URL))
)

if (is.data.frame(steam_data)) {
} else if (is.list(steam_data) && !is.data.frame(steam_data)) {
  steam_data <- bind_rows(steam_data)
}

cat(sprintf("Loaded %d records with %d top-level fields.\n",
            nrow(steam_data), ncol(steam_data)))

glimpse(steam_data[, seq_len(min(15, ncol(steam_data)))])
```

```{r inspect-nulls}
atomic_cols <- names(steam_data)[sapply(steam_data, function(x)
  is.atomic(x) && !is.list(x))]

null_rates <- steam_data[, atomic_cols] |>
  summarise(across(everything(), ~ mean(is.na(.) | . == ""))) |>
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") |>
  arrange(desc(pct_missing)) |>
  filter(pct_missing > 0)

null_rates |>
  slice_head(n = 20) |>
  mutate(pct_missing = percent(pct_missing, accuracy = 0.1)) |>
  knitr::kable(caption = "Top 20 columns by missing-value rate (atomic columns)")
```

------------------------------------------------------------------------
